Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-02-15 ParisEM.tex
Changing results with fair computational time comparison  between PaRIS and FL

\documentclass[12pt,draft]{article}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{makeidx}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{oldgerm}
\usepackage{mathrsfs}
\usepackage[active]{srcltx}
\usepackage{verbatim}
\usepackage[toc,page]{appendix}
\usepackage{aliascnt}
\usepackage{array}
\usepackage{hyperref}
\usepackage[textwidth=4cm,textsize=footnotesize]{todonotes}
\usepackage{xargs}
\usepackage{cellspace}
\usepackage[Symbolsmallscale]{upgreek}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{array}
\geometry{top=3.5cm, bottom=3.5cm, left=3.5cm , right=3.5cm}
\usepackage{fancyhdr}
\pagestyle{fancy}

@@

\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\\
(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}

@@

\section{Numerical experiments}
\label{sec:exp}
%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
%\begin{enumerate}[i)]
%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
%\end{enumerate}
%The performance of the algorithm is investigated on two known models based on diffusion process.

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0. \label{eq:SINE:SDE}
\end{equation}
This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,\ldots,t_{200}=200$ through the observation process $(Y_k)_{0\leq k \leq 200}$:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
\end{equation}
where the $(\varepsilon)_{0\leq k \leq 200}$ are i.i.d. $\mathcal{N}(0,1)$.
In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.

In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50). We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:SINE}.
This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.  Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.

The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag, as the empirical variance of the estimates is the smallest, and the bias is already negligible for $N=400$, whereas small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
%\begin{enumerate}[i)]
%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
%\item a guided bridge process, see \cite{}.
%\end{enumerate}
%\paragraph{Choice of $\mu$}
%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
%\begin{multline*}
%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
%\end{multline*}

@@

The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.

The results for the fixed lag technique are similar to the ones presented in \cite[Figure 1]{olsson:strojby:2011} on the same model. For small lags, the variance of the estimates is small, but the estimation is highly biased. The bias rapidly decreases as the lag increases, together with a  great increase of variance.  PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the variance of the fixed lag when the lag is one. 

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{makeidx}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{oldgerm}
\usepackage{mathrsfs}
\usepackage[active]{srcltx}
\usepackage{verbatim}
\usepackage[toc,page]{appendix}
\usepackage{aliascnt}
\usepackage{array}
\usepackage{hyperref}
\usepackage[textwidth=4cm,textsize=footnotesize]{todonotes}
\usepackage{xargs}
\usepackage{cellspace}
\usepackage[Symbolsmallscale]{upgreek}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{array}
\geometry{top=3.5cm, bottom=3.5cm, left=3.5cm , right=3.5cm}
\usepackage{fancyhdr}
\pagestyle{fancy}

@@

\begin{align}
\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\label{eq:rho:func}
\end{align}

@@

\section{Numerical experiments}
\label{sec:exp}
%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
%\begin{enumerate}[i)]
%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
%\end{enumerate}
%The performance of the algorithm is investigated on two known models based on diffusion process.
We make numerical experiment on two well known SDE models, the sine model and the log-growth model. On both cases, the propagation law for the particle filtering is chosen to be a trade-off between the model's dynamics and the observation, namely, we chose:
$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})$$
where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f. of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, corresponding to the Euler approximation of equation \eqref{eq:target:sde}. In the cases presented here, the resulting distribution is therefore Gaussian with explicit mean and variance.\\
In both cases we compare the performance of the GRand PaRIS algorithm with the fixed lag method of \cite{olsson:strojby:2011}. To achieve a fair comparison, the GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique is achieved using respectively $N=2000$ (for the sine model) and $N=1400$ (for the log growth model), so that both estimations are obtained in a similar computation time.

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0. \label{eq:SINE:SDE}
\end{equation}
This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,\ldots,t_{200}=200$ through the observation process $(Y_k)_{0\leq k \leq 200}$:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
\end{equation}
where the $(\varepsilon)_{0\leq k \leq 200}$ are i.i.d. $\mathcal{N}(0,1)$.
In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.
In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded either globaly on $(x,y)$ or only on $y$, the resulting GRand PaRIS algorithm has therefore a linear complexity.\\
In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50). We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:SINE}.

This estimation is performed 300 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.  Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.

The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag as the bias is already negligible for $N=400$, and with a lower variance than unbiased fixed lag estimates (i.e., in this case, lags larer than 5). Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance. It is worth noting here that the small bias observed when the lag is equal to five is observation dependent, and no general rule can be made on which lag will lead to a neglictible bias. 
%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
%\begin{enumerate}[i)]
%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
%\item a guided bridge process, see \cite{}.
%\end{enumerate}
%\paragraph{Choice of $\mu$}
%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
%\begin{multline*}
%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
%\end{multline*}

@@

The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity.\\
Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.

The results for the fixed lag technique are similar to the ones presented in \cite[Figure 1]{olsson:strojby:2011} on the same model. For small lags, the variance of the estimates is small, but the estimation is highly biased. The bias rapidly decreases as the lag increases, together with a  great increase of variance.  PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the non biased fixed lags estimates. 
